% ══════════════════════════════════════════════════════════════════════════
% §4  The Direct Hammer: SVD
% ══════════════════════════════════════════════════════════════════════════
\section{The Direct Hammer: SVD (The ``Small'' Problem)}
\label{sec:svd}

\subsection{When to Use It}

If your matrix fits comfortably in RAM---say $n \le 10{,}000$---the
\textbf{Singular Value Decomposition} is the most robust solver available.

Every $m \times n$ matrix $A$ (real) admits the factorisation
\[
    A = U \Sigma V^T,
\]
where $U \in \mathbb{R}^{m \times m}$ and $V \in \mathbb{R}^{n \times n}$
are orthogonal, and $\Sigma = \operatorname{diag}(\sigma_1, \dots,
\sigma_{\min(m,n)})$ with $\sigma_1 \ge \sigma_2 \ge \cdots \ge 0$.

\subsection{Why SVD?}

\begin{itemize}[nosep]
    \item \textbf{Rank-deficient systems.}  If $A$ is rank-deficient or
          nearly so, LU and Cholesky will amplify noise.  SVD reveals the
          rank through the singular values and silently discards the
          noise subspace.
    \item \textbf{Rectangular systems.}  Least-squares ($m > n$) and
          minimum-norm ($m < n$) solutions fall out of SVD with no extra
          code.
    \item \textbf{Conditioning at a glance.}  The condition number is
          $\kappa(A) = \sigma_1 / \sigma_{\min}$.  If $\kappa \approx
          1/\varepsilon_{\mathrm{mach}}$, you know immediately that
          \emph{no} algorithm can give you accurate digits.
\end{itemize}

\begin{tipbox}
SVD costs $O(\min(m,n)^2 \cdot \max(m,n))$ flops.  For $n = 10{,}000$
that is $\sim 10^{12}$ flops---a few seconds on a modern CPU.
For $n = 100{,}000$ it becomes minutes; for $n = 10^6$ it is
infeasible.  That is when you switch to sparse/iterative methods
(\S\ref{sec:sparse}--\ref{sec:gmres}).
\end{tipbox}

\subsection{The Eigen One-Liner}

\begin{lstlisting}
// Robust solver for any dense matrix
Eigen::VectorXd x = A.bdcSvd(Eigen::ComputeThinU
                            | Eigen::ComputeThinV).solve(b);
\end{lstlisting}

\texttt{bdcSvd} stands for \emph{Bidiagonal Divide-and-Conquer} SVD,
which is Eigen's fastest dense SVD implementation.
\texttt{ComputeThinU/V} tells it to compute only the first
$\min(m,n)$ columns of $U$ and $V$, saving memory and time.

\subsection{Full Example}

\lstinputlisting[caption={Dense SVD solve with rank and condition number reporting.},
                 label={lst:svd}]
                {code/svd_solve.cpp}
