% ══════════════════════════════════════════════════════════════════════════
% §6  The Iterative Hammer: GMRES
% ══════════════════════════════════════════════════════════════════════════
\section{The Iterative Hammer: GMRES (The ``Big'' Problem)}
\label{sec:gmres}

\subsection{When Direct Solvers Break Down}

Sparse direct solvers (SparseLU, Cholesky) work beautifully up to
$n \sim 10^5$--$10^6$ in 2-D.  In 3-D, however, the factorisation
produces \emph{fill-in}: entries that were zero in $A$ become non-zero
in $L$ and $U$.  For a 3-D mesh with $n$ unknowns, fill-in can grow
as $O(n^{4/3})$, quickly exhausting memory.

Iterative solvers avoid this entirely: they never form $L$ or $U$.
They only need the ability to compute $Ax$ for a given~$x$.

\subsection{How GMRES Works (The 60-Second Version)}

\textbf{Generalised Minimal Residual (GMRES)} builds an orthonormal
basis $\{q_1, q_2, \dots, q_k\}$ for the \emph{Krylov subspace}
\[
    \mathcal{K}_k(A, r_0) = \operatorname{span}\{r_0,\; Ar_0,\;
    A^2 r_0,\; \dots,\; A^{k-1} r_0\},
\]
where $r_0 = b - Ax_0$ is the initial residual.  At step~$k$, it
picks the vector in $x_0 + \mathcal{K}_k$ that minimises the
residual $\|b - Ax\|_2$.

\begin{itemize}[nosep]
    \item \textbf{Convergence:} GMRES converges in at most $n$ steps
          (in exact arithmetic).  In practice it converges much
          sooner if the eigenvalues of $A$ are clustered.
    \item \textbf{Universality:} Unlike Conjugate Gradient (CG),
          GMRES works for \emph{any} non-singular matrix---symmetric
          or not.
    \item \textbf{Memory:} Each iteration stores one new Krylov vector.
          For very large $n$, use \emph{restarted} GMRES($m$),
          which restarts after $m$ vectors.
\end{itemize}

\subsection{Preconditioning: The Secret Sauce}

Without preconditioning, GMRES may need thousands of iterations.
A \textbf{preconditioner} $M \approx A$ transforms the system to
\[
    M^{-1}A\,x = M^{-1}b.
\]
If $M^{-1}A \approx I$, the eigenvalues cluster near~1 and GMRES
converges in very few steps.  The art is choosing $M$ so that:

\begin{enumerate}[nosep]
    \item $M^{-1}v$ is cheap to apply (much cheaper than solving $Ax = v$).
    \item $M^{-1}A$ has clustered eigenvalues.
\end{enumerate}

\begin{definition}[Incomplete LU (ILU)]
Perform LU factorisation of $A$ but \textbf{drop} fill-in entries
below a threshold.  The result $M = \tilde{L}\tilde{U} \approx A$
is cheap to store and apply.  Eigen provides this as
\texttt{IncompleteLUT}.
\end{definition}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Configuration} & \textbf{Typical iterations} & \textbf{Total time} \\
\midrule
No preconditioner      & $1{,}000$--$10{,}000$   & Slow \\
Jacobi (diagonal)      & $500$--$5{,}000$        & Moderate \\
ILU(0)                 & $50$--$200$             & Fast \\
ILU with threshold     & $10$--$50$              & Fastest \\
\bottomrule
\end{tabular}
\end{center}

\begin{warningbox}
A preconditioner that is too aggressive (dropping too many fill-in entries)
can actually \emph{slow down} convergence or cause divergence.
Monitor the residual history; if it stalls or oscillates, tighten the
drop tolerance.
\end{warningbox}

\subsection{GMRES vs.\ BiCGSTAB: Choosing Your Hammer}

Eigen provides two main Krylov solvers.  They are \emph{not}
interchangeable:

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{@{}lcc@{}}
\toprule
& \textbf{GMRES} & \textbf{BiCGSTAB} \\
\midrule
Robustness        & Indestructible      & Can stall or diverge \\
Memory per iter   & $O(n)$ (stores $q_k$) & $O(n)$ (fixed) \\
Iterations/solve  & Monotone decrease   & Non-monotone (``jittery'') \\
Best for          & Any matrix          & Mildly non-symmetric \\
Failure mode      & Slow, never wrong   & Fast, sometimes explodes \\
\bottomrule
\end{tabular}
\end{center}

\begin{warningbox}
If the matrix is \textbf{highly non-symmetric} or has
\textbf{eigenvalues near the origin}, BiCGSTAB can stagnate or
diverge where GMRES would slowly grind to a correct result.
If you need a ``Magic Hammer'' that never fails, use GMRES.
If you have a deadline and can verify convergence, BiCGSTAB is
often faster per iteration.
\end{warningbox}

\subsection{Eigen Implementation}

Eigen ships \texttt{BiCGSTAB} in the core module and \texttt{GMRES}
in the unsupported module (\texttt{<Eigen/IterativeSolvers>}).
Both accept any preconditioner as a template parameter.  The code
below demonstrates all three configurations---unpreconditioned BiCGSTAB,
ILU-preconditioned BiCGSTAB, and ILU-preconditioned GMRES---on the
same problem so you can compare iteration counts and timing directly.

\lstinputlisting[caption={Iterative solve: BiCGSTAB vs.\ GMRES with ILU preconditioning.},
                 label={lst:gmres}]
                {code/iterative_solve.cpp}
